Constructing a universe for the setoid model

### 2020-07-06 paper.tex
changes discussed today

What we would like instead is a formulation of ITT that supports extensionality,
while retaining its convenient computational behaviour.
%
Unfortunately, one can show that canonicity and function extensionality cannot
coexist if propositional equality is given by Martin-L\"of's inductively-defined
identity types.
%
The first step towards a solution is to give up the idea of propositional
equality as a single inductive definition given homogeneously for arbitrary
types. Instead, equality should be \emph{specific} to each type former in the
type theory, or in other words, every type former should be introduced alongside
an explanation of what counts as equality for its elements.
%% In this setting, function extensionality can simply be introduced as the
%% chosen notion of equality on function types.

This idea of pairing types together with their own equality relation goes back
to the notion of \emph{setoid} or \emph{Bishop set}. Setoids provide a quite
natural and useful semantic domain in which to interpret type theory. The first
setoid model was constructed to show consistency of funext for ITT
\cite{hofmann}. This consistency result is implied by the observation that UIP
and funext hold in the setoid model. Moreover, it was shown by Altenkirch
\cite{setoid99} that if the model construction is carried out in a type
theoretic metatheory with a universe of strict (definitionally proof-irrelevant)
propositions, it is possible to define a univalent universe of propositions
satisfying propositional extensionality. The setoid model thus satisfies all the
extensionality principles that we would like to have in type theory. The
question is whether there exists a version of intensional type theory that
supports setoid reasoning, and hence the forms of extensionality enabled by it.

@@

In addition to type formers in $\mType$, we will need the propositional versions
of $\mathbf{1}$, $\Pi$, and $\Sigma$; however these can be defined from their
$\mType$ counterparts via truncation. That is, given $P : \mProp$ and $Q : P \to
\mProp$:
%
\begin{align*}
  \Pi_p\ P\ Q & :\equiv \| \Pi\ (\textsf{Lift} \ P)\ (\textsf{Lift} \circ Q \circ \textsf{unlift}) \| \\
  \Sigma_p\ P\ Q & :\equiv \| \Sigma\ (\textsf{Lift} \ P)\ (\textsf{Lift} \circ Q \circ \textsf{unlift}) \| \\
  \mathbf{1}_p & :\equiv \| \mathbf{1} \|
\end{align*}

@@

Types in a context $\GG$ are given by setoid morphisms from $\GG$ to the
(strict) category of setoids. The definition can be unfolded as follows.

@@

We recall the rules of SeTT that extend regular MLTT below, but with a
variation: whereas the equality types in \cite{mpc19} are stated as elements of
SeTT's internal universe of propositions, here we state the context equalities
as elements of the external, metatheoretic universe $\mProp$. Equality on
dependent types is defined as before in \cite{mpc19}.

@@

Note that $\eqel{\_}{\_}{\_}$ is indexed over equality on the universe, because
$\El$ is a setoid morphism, hence in particular it must respect the setoid
equality of $\setoidU$.
%
We also require data and proofs that make sure we get setoids out of $\El$:
\begin{align*}
  & \textsf{refl}_\El : (A : \setoidU) (x : \El\ A) \to \eqel{\reflu\ A}{x}{x} \\
  & \textsf{sym}_\El : \eqel{p}{x}{x'} \to \eqel{\symu\ p}{x'}{x} \\
  & \textsf{trans}_\El : \eqel{p}{x}{x'} \to \eqel{q}{x'}{x''} \to \eqel{\transu\ p\ q}{x}{x''} \\
  & \coeel : \equ{A}{B} \to \El\ A \to \El\ B \\
  & \cohel : (p : \equ{A}{A'}) \ (x : \El\ A) \to \eqel{p}{x}{\coeel\ p\ x}
\end{align*}
%
We give an inductive definition of $\setoidU$, mutually with a recursive definition
of the 4 functions
$\_\sim_\setoidU\_$, $\textsf{refl}_\setoidU$, $\El$ and $\_\vdash\_\sim_\El\_$.
The other functions are then recursively defined:
\begin{itemize}
\item $\textsf{refl}_\El$ alone,
\item $\textsf{sym}_\setoidU$ and $\textsf{sym}_\El$ mutually,
\item $\textsf{trans}_\setoidU$, $\textsf{trans}_\El$, $\coeel$ and $\cohel$ mutually.
\end{itemize}
The whole construction is quite long, below we only show the more interesting
definitions of $\setoidU$ and $\El$:

@@

We can show that $\Uty$ is closed under $\Pi$ types and bool type, including
$\beta$ and $\eta$ laws. Moreover, all the necessary equations (including
substitution laws) hold definitionally in the metatheory.
The universe can be closed under more constructions if more codes are added to
$\setoidU$.

@@

\begin{align*}
  & \textsf{data}\ \ad{in-U}_0 : \mType \to \mType_1 \\
  & \textsf{data}\ \ad{in-U∼}_0 : \{A\ A' : \mType\} \to (A \to A' \to \mProp) \to \mType_1 \\
  & \textsf{data}\ \ad{in-U}_1 : (A : \mType) \to \ad{in-U}_0\ A \to \mProp_1 \\
  & \textsf{data}\ \ad{in-U∼}_1 : \{A\ A' : \mType\} \to (R : A \to A' \to \mProp) \to \ad{in-U∼}_0\ R \to \mType_1
\end{align*}

@@

Having $\mProp$ and a strong transport principle does help to some extent.
%
However, we would still need extensionality to derive the general eliminators
for $\ad{in-U}$ and $\ad{in-U∼}$. In fact, as explained in the previous section,
to derive the general recursive-recursive eliminators we need to prove that the
correponding graph relations are functional, which can't be done without funext.

Luckily, it turns out that the \emph{simple} elimination principle is sufficient
for our purposes. This eliminator can be defined by simple pattern matching on
the untyped codes, and does not require extensionality or any extra principles
beyond strong transport.

>>>>

What we would like instead is a formulation of ITT that supports extensionality,
while retaining its convenient computational behaviour.
%
Unfortunately, canonicity for Martin-L{\"o}f's inductively defined
identity type says that if two terms are propositionally equal in the
empty context, then they are also definitionally equal. This rules out
function extensionality.
%
The first step towards a solution is to give up the idea of propositional
equality as a single inductive definition given generically for arbitrary
types. Instead, equality should be \emph{specific} to each type former in the
type theory, or in other words, every type former should be introduced alongside
an explanation of what counts as equality for its elements.
%% In this setting, function extensionality can simply be introduced as the
%% chosen notion of equality on function types.

This idea of pairing types together with their own equality relation goes back
to the notion of \emph{setoid} or \emph{Bishop set}. Setoids provide a quite
natural and useful semantic domain in which to interpret type theory. The first
setoid model was constructed to justify function extensionality without relying
on funext in the metatheory \cite{hofmann}. Moreover, it was shown by Altenkirch
\cite{setoid99} that if the model construction is carried out in a type
theoretic metatheory with a universe of strict (definitionally proof-irrelevant)
propositions, it is possible to define a univalent universe of propositions
satisfying propositional extensionality. The setoid model thus satisfies all the
extensionality principles that we would like to have in type theory. The
question is whether there exists a version of intensional type theory that
supports setoid reasoning, and hence the forms of extensionality enabled by it.

@@

In addition to type formers in $\mType$, we will need the propositional versions
of $\mathbf{0}$, $\mathbf{1}$, $\Pi$, and $\Sigma$. The latter three can be defined from their
$\mType$ counterparts via truncation. That is, given $P : \mProp$ and $Q : P \to
\mProp$:
%
\begin{align*}
  \mathbf{1}_p & :\equiv \| \mathbf{1} \| \\
  \Pi_p\ P\ Q & :\equiv \| \Pi\ (\textsf{Lift} \ P)\ (\textsf{Lift} \circ Q \circ \textsf{unlift}) \| \\
  \Sigma_p\ P\ Q & :\equiv \| \Sigma\ (\textsf{Lift} \ P)\ (\textsf{Lift} \circ Q \circ \textsf{unlift}) \|
\end{align*}
We assume that we have
$\mathbf{0}_p:\mProp$ together with $\mathsf{exfalso}_p :\{A:\mType \} \to \mathbf{0}_p \to A$.

@@

Types in a context $\GG$ are given by displayed setoids over $\GG$
with a fibration condition given by $\textsf{coe}$, $\textsf{coh}$.

@@

We recall the rules of SeTT that extend regular MLTT below, but with a
variation: whereas the equality types in \cite{mpc19} are stated as elements of
SeTT's internal universe of propositions, here we state the context equalities
as elements of the external, metatheoretic universe $\mProp$. This makes it easier
to define models of SeTT as contexts do not need to be partially fibrant. Equality on
types is defined as before in \cite{mpc19}.

@@

Note that $\eqel{\_}{\_}{\_}$ is indexed over equality on the universe, because
$\El$ is a displayed setoid over $\setoidU$, hence in particular it must respect the setoid
equality of $\setoidU$.
%
We also require data and proofs that make sure we get setoids out of $\El$:
\begin{align*}
  & \textsf{refl}_\El : (A : \setoidU) (x : \El\ A) \to \eqel{\reflu\ A}{x}{x} \\
  & \textsf{sym}_\El : \eqel{p}{x}{x'} \to \eqel{\symu\ p}{x'}{x} \\
  & \textsf{trans}_\El : \eqel{p}{x}{x'} \to \eqel{q}{x'}{x''} \to \eqel{\transu\ p\ q}{x}{x''} \\
  & \coeel : \equ{A}{B} \to \El\ A \to \El\ B \\
  & \cohel : (p : \equ{A}{A'}) \ (x : \El\ A) \to \eqel{p}{x}{\coeel\ p\ x}
\end{align*}
%
We give an inductive definition of $\setoidU$, mutually with a recursive definition
of the 4 functions
$\_\sim_\setoidU\_$, $\textsf{refl}_\setoidU$, $\El$ and $\_\vdash\_\sim_\El\_$.
The other functions are then recursively defined:
\begin{itemize}
\item $\textsf{refl}_\El$ alone,
\item $\textsf{sym}_\setoidU$ and $\textsf{sym}_\El$ mutually,
\item $\textsf{trans}_\setoidU$, $\textsf{trans}_\El$, $\coeel$ and $\cohel$ mutually.
\end{itemize}
The whole construction is quite long, below we only show the more interesting
definitions of $\setoidU$ and $\El$:

@@

We can show that $\Uty$ is closed under $\Pi$ types and booleans such
that $\EL\,(\mathsf{pi}\,A\,B) \equiv \Pi\,(\EL\,A)\,(\EL\,B)$ and
$\EL\,\mathsf{bool} = \mathsf{Bool}$. The universe can be closed under
more constructions if more codes are added to $\setoidU$.

@@

\begin{align*}
  & \textsf{data}\ \ad{in-U}_0 : \mType \to \mType_1 \\
  & \textsf{data}\ \ad{in-U∼}_0 : \{A\ A' : \mType\} \to (A \to A' \to \mProp) \to \mType_1 \\
  & \textsf{data}\ \ad{in-U}_1 : (A : \mType) \to \ad{in-U}_0\ A \to \mProp_1 \\
  & \textsf{data}\ \ad{in-U∼}_1 : \{A\ A' : \mType\} \to (R : A \to A' \to \mProp) \to \ad{in-U∼}_0\ R \to \mProp_1
\end{align*}

@@

Having $\mProp$ and a strong transport principle does help to some extent.
%
However, we would still need extensionality to derive the general eliminators
for $\ad{in-U}$ and $\ad{in-U∼}$. In fact, as explained in the previous section,
to derive the general recursive-recursive eliminators we need to prove that the
corresponding graph relations are functional, which can't be done without funext.

Luckily, it turns out that the \emph{simple} elimination principle is sufficient
for our purposes: all functions of Section \ref{ii-universe} are defined using simple mutual pattern matching without recursion-recursion.
The simple eliminator can be defined by pattern matching on
the untyped codes, and does not require extensionality or any extra principles
beyond strong transport.
